Asymmetric relationships and their effects on coexistence: Supporting information

### 2023-02-28 paper/SupMat.Rnw
took another look at the Results I edited the other day and made some further presentational improvements
tiny things

%***DAN: I am 90% sure that once I re-read the comment of Ellner saying we do not need to consider really
%big \delta, we will be able to delete the below.
% \section{The $\delta=1$ case}\label{sect:delta1}
% 
% %The i terms
% Setting $\delta=1$ in (\ref{eq:decomp:baseline}) gives 
% \begin{equation}
% \epsilon_i^0 = \mu_i - \mu_j.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:contrib_of_var_E}) gives
% \begin{align}
% \olep_i^E &= \mu_i-\mu_j-\sigma^2/2-\epsilon_i^0 \\
% &= -\sigma^2/2.
% \end{align}
% Setting $\delta=1$ in (\ref{eq:decomp:contrib_of_var_C}) gives
% \begin{align}
% \olep_i^C &= \mu_i - \mu_j + \sigma^2/2 -\epsilon_i^0 \\
% &= \sigma^2/2.
% \end{align}
% Setting $\delta=1$ in (\ref{eq:decomp:last_term_primary}) gives
% \begin{align}
% \olep_i^{EC} &= \mu_i - \mu_j -\olep_i^E - \olep_i^C - \epsilon_i^0 \\
% &= 0.
% \end{align}
% Setting $\delta=1$ in (\ref{eq:decomp:contrib_of_EC_var_per_se}) gives
% \begin{align}
% \olep_i^{(E \shp C)} &= \mu_i - \mu_j -\olep_i^E - \olep_i^C - \epsilon_i^0 \\
% &= 0.
% \end{align}
% Setting $\delta=1$ in (\ref{eq:decomp:storage_effects}) gives
% \begin{align}
% \olep_i^{(EC)} &= \mu_i - \mu_j - (\mu_i -\mu_j) \\
% &= 0.
% \end{align}
% Setting $\delta=1$ in (\ref{eq:decomp:contrib_ATA_2}) gives
% \begin{equation}
% \olep_i^{[E \nat C]}=0.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:contrib_EC}) gives
% \begin{equation}
% \olep_i^{[EC]}=0.
% \end{equation}
% %The j terms
% Setting $\delta=1$ in (\ref{eq:decomp:baseline_j}) gives
% \begin{equation}
% \epsilon_j^0 = 0.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:contrib_of_var_E_j}) gives
% \begin{equation}
% \olep_j^E = -\sigma^2/2.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:contrib_of_var_C_j}) gives
% \begin{equation}
% \olep_j^C = \sigma^2/2.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:j:EC}) gives
% \begin{equation}
% \olep_j^{EC} = 0.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:j:EshpC}) gives
% \begin{equation}
% \olep_j^{(E \shp C)} = 0.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:j:pECp}) gives
% \begin{equation}
% \olep_j^{(EC)} = 0.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:j:EnatC}) gives
% \begin{equation}
% \olep_j^{[E \nat C]} = 0.
% \end{equation}
% Setting $\delta=1$ in (\ref{eq:decomp:j:EChard}) gives
% \begin{equation}
% \olep_j^{[EC]}=0.
% \end{equation}
% 
% For $\delta=1$, the second value of $q_{ij}$ we considered, namely
% $q_{ij}=\frac{\exp(\mu_i)}{(1-\delta)\exp(\mu_j)+\delta \exp(\mu_i)}$, equals $1$, so both the
% options we considered for $q_{ij}$ are $1$ in this case. Combining the above 
% equations using $q_{ij}=1$ gives
% \begin{align}
% \Delta_i^0 &= \mu_i - \mu_j \\
% \Delta_i^E &= 0  \\
% \Delta_i^C &= 0  \\
% \Delta_i^{(E \shp C)} &=  0 \\
% \Delta_i^{[E \nat C]} &= 0  \\
% \Delta_i^{[EC]} &=  0. 
% \end{align}
% Thus the only contribution to coexistence for the $\delta=1$ case is $\Delta_i^0 = \mu_i - \mu_j$,
% which does not depend on $\sigma$ and which is non-positive whenever $\mu_i \leq \mu_j$ (which we assumed,
% without loss of generality).
% So there is no robust coexistence in the $\delta=1$ case for our simple model.

@@

%The Q, K and V functions
Estimates of the temperature dependence functions $Q_i(t)$, $V_i(t)$, and $K_i(t)$, 
produced by earlier authors \cite{Descamps_2005,Ellner_2016,Ellner_2019} and which are part of the parameterization
of the diatom competition model described in the main text, are now described.
\cite{Descamps_2005} estimated values for $Q_i(t)$, $V_i(t)$, and $K_i(t)$ from 
single-species, nine-day batch experiments at four different temperatures ($6^\circ$C, 
$12^\circ$C, $18^\circ$C, $24^\circ$C) (see Table 1 of \cite{Descamps_2005}) and 
used linear interpolations for the
functions $Q_i(t)$, $V_i(t)$, and $K_i(t)$. Their values and functions are 
pictured in Fig. \ref{fig:VKQ} using open circles and dashed lines.
\cite{Ellner_2016, Ellner_2019} used slightly modified values and functions,
pictured in Fig. \ref{fig:VKQ} by filled circles and solid lines. 
Our investigation follows \cite{Ellner_2016, Ellner_2019} and 
uses their modified functions.

%Defintions of E and C
Our decomposition of GWR is defined by the choices $E_i(t)=\theta(t)$, 
$C_{i}(t)=(K_i(t)+S(t))/S(t)$, which then imply $r_i(E_i,C_{i})=\frac{V_i(E_i)}{C_{i}} -D$. 
This choice follows one of the choices made in the R computer code provided by \citep{Ellner_2016,Ellner_2019}.
Other choices are possible. For instance, in addition to this choice, 
\cite{Ellner_2016} also considers an alternative choice $E_i(t)=V_i(t)$,
as well as some other possibilities discussed in their supporting materials and 
represented in their computer code.

%Outline how the quantities E_i and C_i\i were obtained from simulations. 
To obtain $E_i$ and $C_{i \bs i}$, the competition model was simulated for $T=3000$ time steps.
Environmental temperature fluctuations were defined by the paramters $a$, $P$, and $\theta_0$ using 
$\theta(t)=asin(2\pi t/P)+\theta_0$, and this determined $E_i(t)$ for $i=1,2$. To obtain 
$C_{i \bs i}$, a simulation was carried out starting from $x_i=0$ and $x_j=20$.
Dilution was set to $D=0.09$ and inflow silicate concentration was set at $S_0=35$. A burn in time 
of $\frac{2}{3}T$ was cut from the simulation and the $S(t)$ time series from the remainder of the simulation,
along with $K_i(t)=K_i(\theta(t))$ were plugged in to the expression $C_{i}(t)=(K_i(t)+S(t))/S(t)$.

@@

The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example explored for goal G4 of the main text (see also SI section \ref{sect:diatom_model}), 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combined to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.

>>>>

@@

%The Q, K and V functions
Estimates of the temperature dependence functions $Q_i(t)$, $V_i(t)$, and $K_i(t)$, 
produced by earlier authors \citep{Descamps_2005,Ellner_2016,Ellner_2019} and which are part of the parameterization
of the diatom competition model described in the main text, are now described.
\cite{Descamps_2005} estimated values for $Q_i(t)$, $V_i(t)$, and $K_i(t)$ from 
single-species, nine-day batch experiments at four different temperatures ($6^\circ$C, 
$12^\circ$C, $18^\circ$C, $24^\circ$C) (see Table 1 of \citealp{Descamps_2005}) and 
used linear interpolations for the
functions $Q_i(t)$, $V_i(t)$, and $K_i(t)$. Their values and functions are 
pictured in Fig. \ref{fig:VKQ} using open circles and dashed lines.
\cite{Ellner_2016, Ellner_2019} used slightly modified values and functions,
pictured in Fig. \ref{fig:VKQ} by filled circles and solid lines. 
Our investigation follows \cite{Ellner_2016, Ellner_2019} and 
uses their modified functions.

%Defintions of E and C
Our decomposition of GWR is defined by the choices $E_i(t)=\theta(t)$, 
$C_{i}(t)=(K_i(t)+S(t))/S(t)$, which then imply $r_i(E_i,C_{i})=\frac{V_i(E_i)}{C_{i}} -D$. 
This choice follows one of the choices made in the R computer code provided by \cite{Ellner_2016} and 
\cite{Ellner_2019}.
Other choices are possible. For instance, in addition to this choice, 
\cite{Ellner_2016} also considers an alternative choice $E_i(t)=V_i(t)$,
as well as some other possibilities discussed in their supporting materials and 
represented in their computer code.

%Outline how the quantities E_i and C_i\i were obtained from simulations. 
To obtain $E_i$ and $C_{i \bs i}$, the competition model was simulated for $T=3000$ time steps.
Environmental temperature fluctuations were defined by the paramters $a$, $P$, and $\theta_0$ using 
$\theta(t)=a\sin(2\pi t/P)+\theta_0$, and this determined $E_i(t)$ for $i=1,2$. To obtain 
$C_{i \bs i}$, a simulation was carried out starting from $x_i=0$ and $x_j=20$.
Dilution was set to $D=0.09$ and inflow silicate concentration was set at $S_0=35$. A burn in time 
of $\frac{2}{3}T$ was cut from the simulation and the $S(t)$ time series from the remainder of the simulation,
along with $K_i(t)=K_i(\theta(t))$ were plugged in to the expression $C_{i}(t)=(K_i(t)+S(t))/S(t)$.

@@

The decomposition of storage effects and quantification of contributions of ATAs depends on 
removing asymmetric tail associations from the bivariate random variable
$(E_i,C_{i \bs i})$, while maintaining its correlation and marginal distributions, to produce a new
random variable $(E_i^\nat,C_{i \bs i}^\nat)$.
In the Theory section of the main text, we decribed this removal process
when the distributions of $E_i$ and $C_{i \bs i}$ are known exactly, but for the 
diatom example, 
we only had samples from $(E_i,C_{i \bs i})$, based on simulations. We here describe how to
produce a sample from $(E_i^\nat,C_{i \bs i}^\nat)$ in this modified context. 
Let $(E_{i,k},C_{i \bs i,k})$, $k=1,\ldots,T$ denote the samples from $(E_i,C_{i \bs i})$.
Let $\rk(E_{i,k})$ be the rank of $E_{i,k}$ in the set 
of $E_{i,k}$, $k=1,\ldots,T$, where the smallest value has rank $1$. Likewise 
define $\rk(C_{i \bs i,k})$ as the rank of $C_{i \bs i,k}$ in the set
of $C_{i \bs i,k}$, $k=1,\ldots,T$. Define the ``normalized rank''
$\nrk(E_{i,k})$ as $\rk(E_{i,k})/(T+1)$, and likewise define 
$\nrk(C_{i \bs i,k})$ as $\rk(C_{i \bs i,k})/(T+1)$. Letting $\varphi$ be the CDF
of a standard normal distribution, the samples $\varphi^{-1}(\nrk(E_{i,k}))$, $k=1,\ldots,T$
are standard normally distributed, as are the samples 
$\varphi^{-1}(\nrk(C_{i \bs i,k}))$, $k=1,\ldots,T$. Let $\rho_i$ be the covariance 
of these samples. We then define a bivariate normal random variable 
$(e_i,c_{i \bs i})$ such that $\E(e_i)=\E(c_{i \bs i})=0$, $\var(e_i)=\var(c_{i \bs i})=1$,
and we denote $\cov(e_i,c_{i \bs i})$ by $\ddot{\rho}_i$. We used $\ddot{\rho}_i = \rho_i$.
Generating a sample $(e_{i,k},c_{i \bs i,k})$, $k=1,\ldots,T$ from this distribution,
we permute the ordered set $(E_{i,1},E_{i,2},\ldots,E_{i,T})$ so that the ranks of the 
permuted ordered set equal the ranks of the corresponding elements of the ordered
set $(e_{i,1},e_{i,2},\ldots,e_{i,T})$. We likewise permute the ordered set 
$(C_{i \bs i,1},C_{i \bs i,2},\ldots,C_{i \bs i,T})$ so the ranks of the permuted result
match the ranks of $(c_{i \bs i,1},c_{i \bs i,2},\ldots,c_{i \bs i,T})$.
This step works as long as there are no ties among the $E_{i,k}$ or among
the $C_{i \bs i,k}$. There were no ties when we applied these methods to the diatom system. 
The two permuted ordered sets, here denoted $(E_{i,1}^\nat,E_{i,2}^\nat,\ldots,E_{i,T}^\nat)$
and $(C_{i \bs i,1}^\nat,C_{i \bs i,2}^\nat,\ldots,C_{i \bs i,T}^\nat)$,
when combined to make ordered pairs, form the desired sample from 
$(E_i^\nat,C_{i \bs i}^\nat)$. Because $(E_{i,1}^\nat,\ldots,E_{i,T}^\nat)$
is a permutation of $(E_{i,1},\ldots,E_{i,T})$ and 
$(C_{i \bs i,1}^\nat,\ldots,C_{i \bs i,T}^\nat)$ is a permutation of
$(C_{i \bs i,1},\ldots,C_{i \bs i,T})$, marginal distributions of our
sample from $(E_i^\nat,C_{i \bs i}^\nat)$ were the same as those of our original 
sample from $(E_i, C_{i \bs i})$. The choice of bivariate normal
$(e_i,c_{i \bs i})$ ensures that our resulting sample shows symmetric tail association.
Finally, we caution that sampling variation means any quantity computed from our resulting sample 
$(E_{i,k}^\nat,C_{i \bs i,k}^\nat)$, $k=1,\ldots,T$ should be recomputed for 
multiple such samples, all computed via the same process, and a median of 
the resulting distribution of values should be used.
